quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-11-10 paper-quollr.Rmd
Checking wordings

- `hex_bin_obj`: the object created by `hex_binning()`, which inlcudes hex binning configurations. It includes the binwidth (`a1`), binheight (`a2`), the number of bins along each axis (`b1`, `b2`), the centroids of all hexagons, polygon coordinates, and the assignment of each data point to the hexagon.

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. To begin the workflow, users need two inputs: the high-dimensional dataset and the corresponding nonlinear dimensionality reduction (NLDR) layout. The high-dimensional data must contain a unique `ID` column, with data columns prefixed by the letter `"x"` (e.g., `x1`, `x2`, etc.). The NLDR dataset should include embedding coordinates labeled as `emb1` and `emb2`, ensuring one-to-one correspondence with the high-dimensional data through the shared `ID`.

@@

For this example, we use the UMAP layout, which is produced using `n_neighbors =` $46$ and `min_dist =` $0.9$.

@@

The model generation involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, (v) removing low-density hexagons, and (vi) lifting the model to high-dimensions.

@@

The function `calc_bins_y()` determines the configuration of the hexagonal grid by computing the number of bins along the y-axis (`b2`), the hexagon width (`a1`), and height (`a2`). This function accepts (1) an object (`nldr_scaled_obj`) containing three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); (2) the number of bins along the x-axis (`b1`), and (3) the buffer amount as a proportion (`q`). The buffer ensures that the grid fully covers the data space by extending one hexagon width ($a_1$) and height ($a_2$) beyond the observed data in all directions. By default, $q = 0.1$, but it must be set to a value smaller than the minimum data value to avoid exceeding the data range.

>>>>

- `hex_bin_obj`: the object created by `hex_binning()`, which includes hexagon grid configurations. It includes the binwidth (`a1`), binheight (`a2`), the number of bins along each axis (`b1`, `b2`), the centroids of all hexagons, polygon coordinates, and the assignment of each data point to the hexagon.

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. To begin the workflow, user needs two inputs: the high-dimensional dataset and the corresponding NLDR data. The high-dimensional data must contain a unique `ID` column, with data columns prefixed by the letter `"x"` (e.g., `x1`, `x2`, etc.). The NLDR dataset should include embedding coordinates labeled as `emb1` and `emb2`, ensuring one-to-one correspondence with the high-dimensional data through the shared `ID`.

@@

For this example, we use the UMAP layout, which is used `n_neighbors =` $46$ and `min_dist =` $0.9$ as hyper-parameters.

@@

@@

The function `calc_bins_y()` determines the configuration of the hexagon grid by computing the number of bins along the y-axis (`b2`), the hexagon binwidth (`a1`), and binheight (`a2`). This function accepts (i) an object (`nldr_scaled_obj`) containing three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); (ii) the number of bins along the x-axis (`b1`), and (iii) the buffer amount as a proportion (`q`). The buffer ensures that the grid fully covers the data space by extending one hexagon width ($a_1$) and height ($a_2$) beyond the observed data in all directions. By default, $q = 0.1$, but it must be set to a value smaller than the minimum data value to avoid exceeding the data range.

### 2025-11-10 paper-quollr.tex
Checking wordings

\begin{itemize}
\item
  \texttt{scaled\_nldr\_obj}: the output of \texttt{gen\_scaled\_data()}, which rescales the embedding to the range \([0, 1] \times [0, y_{2,\text{max}}]\), where \(y_{2,\text{max}} = r_2 / r_1\) is the ratio of the embedding ranges. It includes the scaled coordinates (\texttt{scaled\_nldr}) and the original limits (\texttt{lim1}, \texttt{lim2}).
\item
  \texttt{hex\_bin\_obj}: the object created by \texttt{hex\_binning()}, which inlcudes hex binning configurations. It includes the binwidth (\texttt{a1}), binheight (\texttt{a2}), the number of bins along each axis (\texttt{b1}, \texttt{b2}), the centroids of all hexagons, polygon coordinates, and the assignment of each data point to the hexagon.
\item
  \texttt{highd\_vis\_model}: the main model object returned by \texttt{fit\_highd\_model()}. It stores all components of the fitted model, including the scaled NLDR data (\texttt{nldr\_scaled\_obj}), the hexagonal bin configurations (\texttt{hb\_obj}), the averaged \(p\text{-}D\) summaries for each bin (\texttt{model\_highd}), the corresponding \(2\text{-}D\) bin centroids (\texttt{model\_2d}), and the triangulated mesh connecting neighboring bins (\texttt{trimesh\_data}).
\end{itemize}

@@

The following demonstration of the package's functionality assumes \texttt{quollr} has been loaded. To begin the workflow, users need two inputs: the high-dimensional dataset and the corresponding nonlinear dimensionality reduction (NLDR) layout. The high-dimensional data must contain a unique \texttt{ID} column, with data columns prefixed by the letter \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, etc.). The NLDR dataset should include embedding coordinates labeled as \texttt{emb1} and \texttt{emb2}, ensuring one-to-one correspondence with the high-dimensional data through the shared \texttt{ID}.

@@

For this example, we use the UMAP layout, which is produced using \texttt{n\_neighbors\ =} \(46\) and \texttt{min\_dist\ =} \(0.9\).

@@

The function \texttt{calc\_bins\_y()} determines the configuration of the hexagonal grid by computing the number of bins along the y-axis (\texttt{b2}), the hexagon width (\texttt{a1}), and height (\texttt{a2}). This function accepts (1) an object (\texttt{nldr\_scaled\_obj}) containing three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); (2) the number of bins along the x-axis (\texttt{b1}), and (3) the buffer amount as a proportion (\texttt{q}). The buffer ensures that the grid fully covers the data space by extending one hexagon width (\(a_1\)) and height (\(a_2\)) beyond the observed data in all directions. By default, \(q = 0.1\), but it must be set to a value smaller than the minimum data value to avoid exceeding the data range.

>>>>

\begin{itemize}
\item
  \texttt{scaled\_nldr\_obj}: the output of \texttt{gen\_scaled\_data()}, which rescales the embedding to the range \([0, 1] \times [0, y_{2,\text{max}}]\), where \(y_{2,\text{max}} = r_2 / r_1\) is the ratio of the embedding ranges. It includes the scaled coordinates (\texttt{scaled\_nldr}) and the original limits (\texttt{lim1}, \texttt{lim2}).
\item
  \texttt{hex\_bin\_obj}: the object created by \texttt{hex\_binning()}, which includes hexagon grid configurations. It includes the binwidth (\texttt{a1}), binheight (\texttt{a2}), the number of bins along each axis (\texttt{b1}, \texttt{b2}), the centroids of all hexagons, polygon coordinates, and the assignment of each data point to the hexagon.
\item
  \texttt{highd\_vis\_model}: the main model object returned by \texttt{fit\_highd\_model()}. It stores all components of the fitted model, including the scaled NLDR data (\texttt{nldr\_scaled\_obj}), the hexagonal bin configurations (\texttt{hb\_obj}), the averaged \(p\text{-}D\) summaries for each bin (\texttt{model\_highd}), the corresponding \(2\text{-}D\) bin centroids (\texttt{model\_2d}), and the triangulated mesh connecting neighboring bins (\texttt{trimesh\_data}).
\end{itemize}

@@

The following demonstration of the package's functionality assumes \texttt{quollr} has been loaded. To begin the workflow, user needs two inputs: the high-dimensional dataset and the corresponding NLDR data. The high-dimensional data must contain a unique \texttt{ID} column, with data columns prefixed by the letter \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, etc.). The NLDR dataset should include embedding coordinates labeled as \texttt{emb1} and \texttt{emb2}, ensuring one-to-one correspondence with the high-dimensional data through the shared \texttt{ID}.

@@

For this example, we use the UMAP layout, which is used \texttt{n\_neighbors\ =} \(46\) and \texttt{min\_dist\ =} \(0.9\) as hyper-parameters.

@@

The function \texttt{calc\_bins\_y()} determines the configuration of the hexagon grid by computing the number of bins along the y-axis (\texttt{b2}), the hexagon binwidth (\texttt{a1}), and binheight (\texttt{a2}). This function accepts (i) an object (\texttt{nldr\_scaled\_obj}) containing three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); (ii) the number of bins along the x-axis (\texttt{b1}), and (iii) the buffer amount as a proportion (\texttt{q}). The buffer ensures that the grid fully covers the data space by extending one hexagon width (\(a_1\)) and height (\(a_2\)) beyond the observed data in all directions. By default, \(q = 0.1\), but it must be set to a value smaller than the minimum data value to avoid exceeding the data range.
